Modeling the Linux page cache for accurate simulation of data-intensive applications

### 2020-12-11 paper_io_sim.tex
Update some text according to comments

    \section{Methods}
    \label{method}
    We separate our simulation model in two components, the I/O
    Controller and the Memory Manager, which together simulate
    file reads and writes (Figure~\ref{fig:interaction}).
    To read or write a file chunk, a simulated application sends a
    request to the I/O Controller. The I/O Controller interacts as needed with
    the Memory Manager to free memory through flushing or eviction,
    and to read or write cached data. The Memory Manager
    implements these operations, simulates periodical flushing
    and eviction, and reads or writes to disk when necessary.
    In case the writethrough strategy is used, the I/O Controller directly
    writes to disk \tristan{Then how is the written data added to cache?}.

@@

    \begin{algorithm}\caption{Periodical flushing simulation in I/O Controller}\label{alg:pdflush}
        \small
        \begin{algorithmic}[1]
            \Input
                \Desc{in}{page cache inactive list}
                \Desc{ac}{page cache active list}
                \Desc{t}{predefined flushing time interval}
                \Desc{exp}{predefined expiration time}
                \Desc{sm}{storage simulation model}
               \EndInput
               \While{host is on}
                \State blocks = expired\_blocks(exp, in) + expired\_blocks(exp, ac)
                \State flushing\_time = sm.write(blocks)
                \State for blk in blocks: blk.dirty = 0 \tristan{with this logic a flushed block can remain dirty for a long time. Is it what you implemented?
                it would be wise to set the dirty flag to 0 right after the block is written}
                \If{flushing\_time $<$ t}
                    \State sleep(t - flushing\_time)
                \EndIf  % Else: do nothing, continue the loop
            \EndWhile
        \end{algorithmic}
    \end{algorithm}

@@

        The real application is a workflow of the Nighres
        toolbox~\cite{huntenburg2018nighres} that implements skull
        stripping in high-resolution brain images, an operation common to
        many neuroimage analysis pipelines. The workflow is implemented as a Python
        script that calls Java image-processing routines. We used Python
        3.6, Java 8, and Nighres \tristan{add version}. We patched the
        skull stripping workflow to monitor task CPU times and inject them
        in the simulation, to focus our analysis on I/O times \tristan{add
        link to patched code}.

@@

            \begin{table}[b]
            \centering
            \begin{tabularx}{0.8\columnwidth}{c>{\centering\arraybackslash}X}
            \toprule
                Input size (GB)  & CPU time (second)\\
            \midrule
                3      & 4.4 \\
                20  & 28 \\
                50  & 75 \\
                75  & 110 \\
                100  & 155 \\
            \bottomrule
            \end{tabularx}
            \caption{Task CPU times used in the simulation. \tristan{are the input files in GB or in GiB? If GiB, then change the abbreviation everywhere.}
            \tristan{add nighres CPU times when available}}
            \label{table:cputime}
            \end{table}

@@

            Our fourth and final experiment (\textit{Exp Nighres})
            simulates the Nighres skull stripping workflow
            \tristan{describe data, and if the execution was single node or
            distributed.}

@@

        Simulated memory profiles are highly consistent with the real ones
        (Figure~\ref{fig:single_memprof}). With 20~GB files, memory
        profiles almost exactly match the real ones, although dirty data
        seems to be flushing faster in real life than in simulation
        \tristan{any idea why?}, a behavior also observed with 100~GB files. With
        100~GB files, used memory reaches total memory during the first write,
        triggering dirty data flushing. Used memory drops to cached memory
        when application tasks release anonymous memory. Simulated cached
        memory is highly consistent with real values, except toward the end
        of Read 3 where it slightly increases in simulation but not in
        reality \tristan{why?}. As expected, dirty data remains under the
        dirty ratio. The Python prototype and \wrench-cache exhibit nearly
        identical memory profiles, which reinforces the confidence in our
        implementation.

        The content of the simulated memory cache is also highly consistent
        with reality (Figure~\ref{fig:single_cache}). With 20~GB files, the
        simulated cache content exactly matches reality, since all files fit in page cache. With 100~GB files, a slight
        discrepancy is observed after Write 2, which explains the
        simulation error previously mentioned in Read 3. In the real execution indeed,
        File 3 is entirely cached after Write 2, whereas in the simulated execution, part
        of it is on disk. This is due to the fact that the simulated flushing and eviction
        was triggered after File 3 was already written, while in reality these operations
        happened before File 3 was written \tristan{Dzung, pls check that. Also, why doesn't it happen irl?}.

>>>>

    \section{Methods}
    \label{method}
    We separate our simulation model in two components, the I/O
    Controller and the Memory Manager, which together simulate
    file reads and writes (Figure~\ref{fig:interaction}).
    To read or write a file chunk, a simulated application sends a
    request to the I/O Controller. The I/O Controller interacts as needed with
    the Memory Manager to free memory through flushing or eviction,
    and to read or write cached data. The Memory Manager
    implements these operations, simulates periodical flushing
    and eviction, and reads or writes to disk when necessary.
    In case the writethrough strategy is used, the I/O Controller directly writes to disk, 
    cache is flushed if needed and written data is added to page cache.

@@

    \begin{algorithm}\caption{Periodical flushing simulation in I/O Controller}\label{alg:pdflush}
        \small
        \begin{algorithmic}[1]
            \Input
                \Desc{in}{page cache inactive list}
                \Desc{ac}{page cache active list}
                \Desc{t}{predefined flushing time interval}
                \Desc{exp}{predefined expiration time}
                \Desc{sm}{storage simulation model}
               \EndInput
               \While{host is on}
                \State blocks = expired\_blocks(exp, in) + expired\_blocks(exp, ac)
                \State flushing\_time = 0
                \For{blk in blocks} 
                  \State blk.dirty = 0 
                  \State flushing\_time = flushing\_time + sm.write(blocks)
                \EndFor
                \If{flushing\_time $<$ t}
                    \State sleep(t - flushing\_time)
                \EndIf  
            \EndWhile
        \end{algorithmic}
    \end{algorithm}

@@

        The real application is a workflow of the Nighres
        toolbox~\cite{huntenburg2018nighres} that implements skull
        stripping in high-resolution brain images, an operation common to
        many neuroimage analysis pipelines. The workflow is implemented as a Python
        script that calls Java image-processing routines. We used Python
        3.6, Java 8, and Nighres 1.3.0. We patched the
        skull stripping workflow to monitor task CPU times and inject them
        in the simulation, to focus our analysis on I/O times, the patched code is 
        available at \url{https://github.com/dohoangdzung/nighres}.

@@

            \begin{table}[b]
            \centering
            \begin{tabularx}{0.8\columnwidth}{c>{\centering\arraybackslash}X}
            \toprule
                Input size (GB)  & CPU time (second)\\
            \midrule
                3      & 4.4 \\
                20  & 28 \\
                50  & 75 \\
                75  & 110 \\
                100  & 155 \\
            \bottomrule
            \end{tabularx}
            \caption{Task CPU times used in the simulation.
%            \tristan{are the input files in GB or in GiB? If GiB, then change the abbreviation everywhere.}
%  Dzung: it's in GB
            \tristan{add nighres CPU times when available}}
            \label{table:cputime}
            \end{table}

@@

            Our fourth and final experiment (\textit{Exp Nighres})
            simulates the Nighres CRUISE cortical reconstruction workflow
            \tristan{describe data, and if the execution was single node or
            distributed.}

@@

        Simulated memory profiles are highly consistent with the real ones
        (Figure~\ref{fig:single_memprof}). With 20~GB files, memory
        profiles almost exactly match the real ones, although dirty data
        seems to be flushing faster in real life than in simulation
        \tristan{any idea why?}, a behavior also observed with 100~GB files. With
        100~GB files, used memory reaches total memory during the first write,
        triggering dirty data flushing. Used memory drops to cached memory
        when application tasks release anonymous memory. Simulated cached
        memory is highly consistent with real values, except toward the end
        of Read 3 where it slightly increases in simulation but not in
        reality. Our explanation could be that we allocate anonymous memory 
        chunk-by-chunk to application, while anonymous memory is allocated 
        when as soon as task starts.
        As expected, dirty data remains under the
        dirty ratio. The Python prototype and \wrench-cache exhibit nearly
        identical memory profiles, which reinforces the confidence in our
        implementation.

        The content of the simulated memory cache is also highly consistent
        with reality (Figure~\ref{fig:single_cache}). With 20~GB files, the
        simulated cache content exactly matches reality, since all files fit in page cache. With 100~GB files, a slight
        discrepancy is observed after Write 2, which explains the
        simulation error previously mentioned in Read 3. In the real execution indeed,
        File 3 is entirely cached after Write 2, whereas in the simulated execution, 
        only a part of it is cached. 
        This is due to the fact that the in simulation, data is flushed and evicted 
        from the inactive list first, then from the active list. Thus, File 2, which is 
        accessed again by Read 2 and is cached in the active list, is not flushed and evicted, 
        while File 3 is flushed and evicted since it is accessed once only and cached 
        in the inactive list. This may not correctly simulate what really happens 
        in reality.
